---
title: "Exercises 2. Matrices"
author: "Tiffany Cheng"
output: pdf_document
---

## 1. Suppose

$$ A = \begin{bmatrix}
1 & 1 & 3\\ 
5 & 2 & 6\\ 
-2 & -1 & -3
\end{bmatrix}$$
(a) Check that $A^3 = 0$ where 0 is a 3 $\times$ 3 matrix with every entry equal to 0.
```{r}
A <- matrix(c(1,1,3,5,2,6,-2,-1,-3), nrow = 3, byrow = TRUE)
A%*%A%*%A
```

(b) Replace the third column of A by the sum of the second and third columns.
```{r}
A[,3] <- A[,2]+A[,3]
A
```

******

## 2. Create the following matrix B with 15 rows

$$B = \begin{bmatrix}
10 & -10 & 10\\
10 &  -10 & 10\\
 \cdots & \cdots  & \cdots \\
10 & -10 & 10
\end{bmatrix} $$

Calculate the 3x3 matrix $B^TB$. (Look at the help for `crossprod`.)
```{r}
B <- matrix(c(rep(10,15),rep(-10,15),rep(10,15)), nrow = 15)
crossprod(B)
```

******

## 3. Create a 6 x 6 matrix matE with every element equal to 0. 
Check what the functions `row` and `col` return when applied to `matE`. Hence create the 6 $\times$ 6 matrix:

$$\begin{bmatrix}
0 &1  & 0 & 0 & 0 & 0\\
1 &0 & 1 &0  & 0 &0 \\
0 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 0\\
0& 0 & 0 & 1 & 0 & 1\\
0& 0 & 0 & 0 & 1 & 0
\end{bmatrix} $$  

```{r}
matE <- matrix(rep(0,36), nrow = 6, byrow = TRUE)
matE[abs(row(matE)-col(matE))==1] <- 1
matE
```

******

## 4. Look at the help for the function outer(). Hence, create the following patterned matrix:

$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 & 2 & 3 & 4 & 5\\
2 & 3 & 4 & 5 & 6\\
3 & 4 & 5 & 6 & 7\\
4 & 5 & 6 & 7 & 8
\end{bmatrix} $$

```{r}
a <- 0:4
A <- outer(a,a,"+")
A
```

******

## 5. Create the following patterned matrices. In each case, your solution should make use of the special form of the matrix—this means that the solution should easily generalise to creating a larger matrix with the same structure and should not involve typing in all the entries in the matrix.

(a)
$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 &2 & 3 & 4 & 0\\
2 & 3 & 4 & 0 & 1\\
3 & 4 & 0 & 1 & 2\\
4 & 0 & 1 & 2 & 3
\end{bmatrix}  $$
```{r}
a <- 0:4
A <- outer(a,a,"+")%%5
A
```

(b)
$$ \begin{bmatrix}
0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9\\
1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 0\\
\vdots  & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
8 & 9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8
\end{bmatrix} $$
```{r}
b <- 0:9
B <- outer(b,b,"+")%%10
B
```

(c)
$$\begin{bmatrix}
0 & 8 & 7 & 6 & 5 & 4 &3  & 2 & 1\\
1 & 0 & 8 & 7 & 6 & 5 & 4 & 3 & 2\\
2 & 1 & 0 & 8 & 7 & 6 & 5 & 4 & 3\\
3 & 2 & 1 & 0 & 8 & 7 & 6 & 5 & 4\\
4 & 3 & 2 & 1 & 0 & 8 & 7 & 6 & 5\\
5 & 4 & 3 & 2 & 1 & 0 & 8 & 7 & 6\\
6 & 5 & 4 & 3 & 2 & 1 & 0 & 8 & 7\\
7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 & 8\\
8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0
\end{bmatrix} $$
```{r}
c <- 0:8
C <- outer(c,c,"-")%%9
C
```

******

## 6. Solve the following system of linear equations in five unknowns by considering an appropriate matrix equation Ax = y.

$x_1 + 2x_2 + 3x_3 + 4x_4 + 5x_5 =  7$  
$2x_1 +  x_2 + 2x_3 + 3x_4 + 4x_5 = -1$  
$3x_1 + 2x_2 +  x_3 + 2x_4 + 3x_5 = -3$  
$4x_1 + 3x_2 + 2x_3 +  x_4 + 2x_5 =  5$  
$5x_1 + 4x_2 + 3x_3 + 2x_4 +  x_5 = 17$  

Make use of the special form of the matrix A. The method used for the solution should easily generalize to a larger set of equations where the matrix A has the same structure; hence the solution should not involve typing in every number of A.
```{r}
Y <- c(7,-1,-3,5,17)
A <- matrix(rep(0,25), nrow = 5, byrow = TRUE)
A <- abs(row(A)-col(A))+1
A
X <- solve(A,Y)
X
```

******

## 7. Create a 6 x 10 matrix of random integers chosen from 1,2,...,10 by executing the following two lines of code:
```{r}
set.seed(75)  
aMat <- matrix(sample(10, size=60, replace=TRUE), nr=6)
aMat
```

(a) Find the number of entries in each row which are greater than 4.
```{r}
sum(aMat>4)
```

(b) Which rows contain exactly two occurrences of the number seven?
```{r}
which(rowSums(aMat==7)==2)
```

(c) Find those pairs of columns whose total (over both columns) is greater than 75. The answer should be a matrix with two columns; so, for example, the row (1,2) in the output matrix means that the sum of columns 1 and 2 in the original matrix is greater than 75. Repeating a column is permitted; so, for example, the final output matrix could contain the rows (1,2), (2,1), and (2,2).
What if repetitions are not permitted? Then only (1,2) from (1,2),(2,1) and (2,2) would be permitted.
```{r}
sumtwocolumns <- colSums(aMat)
which(outer(sumtwocolumns,sumtwocolumns,"+")>75,arr.ind = TRUE)
```

******

## 8. Calculate

(a) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+j)}$
```{r}
sum((1:20)^4)*sum(1/(3+(1:5)))
```

(b) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+ij)}$
```{r}
sum((1:20)^4/(3+outer(1:20,1:5,"*")))
```

(c) $\sum\limits_{i=1}^{10} \sum\limits_{j=1}^i \frac{i^4}{(3+ij)}$
```{r}
sum((1:10)^4/(3+outer(1:10,1:10,"*")))
```
